---
title: "Assignment 5: Data Visualization"
author: "Aurora McCollum"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 

getwd()
library(tidyverse)
library(cowplot)
library(lubridate)


PeterPaul.chem.nutrients <- read.csv(
  "/Users/rorymccollum/Desktop/Rdata/Environmental_Data_Analytics_2022/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", 
  stringsAsFactors = TRUE)

Niwot.litter <- read.csv(
  "/Users/rorymccollum/Desktop/Rdata/Environmental_Data_Analytics_2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
  stringsAsFactors = TRUE)

#2 

class(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients$sampledate<-ymd(PeterPaul.chem.nutrients$sampledate)
class(PeterPaul.chem.nutrients$sampledate)

class(Niwot.litter$collectDate)
Niwot.litter$collectDate<-ymd(Niwot.litter$collectDate)
class(Niwot.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3

rorystheme<-theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        panel.background = element_rect(fill = "blanchedalmond"),
        legend.position = "right")

theme_set(rorystheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#4

phorus_by_phate<-ggplot(PeterPaul.chem.nutrients, aes(x=tp_ug, y=po4, color=lakename))+
   xlim(0,155)+ ylim(0, 50)+
  xlab("Total Phosphorus") + ylab("Phosphate")+
  labs(color="Lake Name")+
  geom_point(alpha=0.6)+
  geom_smooth(method = lm, color="black")
  
print(phorus_by_phate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#5

#Temperature
temp<-ggplot(PeterPaul.chem.nutrients)+
  geom_boxplot(aes(x=month, y=temperature_C, color=lakename))+
  ylab("Temperature (C)") + xlab("Month")+
  labs(color="Lake Name")
  
print(temp)


#TP
TP<-ggplot(PeterPaul.chem.nutrients)+
  geom_boxplot(aes(x=month, y=tp_ug, color=lakename))+
  ylab("Total Phosphorus") + xlab("Month")+
  labs(color="Lake Name")
  
print(TP)

#TN
TN<-ggplot(PeterPaul.chem.nutrients)+
  geom_boxplot(aes(x=month, y=tn_ug, color=lakename))+
  ylab("Total Nitrogen") + xlab("Month")+
  labs(color="Lake Name")
  
print(TN)

#all together

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
